---
title: "colondata"
output: html_document
date: "2024-11-26"
---

```{r}
library(survival)

colon <- survival::colon
colnames(colon)
```

Data Preparation

```{r}
library(dplyr)

# Filter death
colon <- as_tibble(colon)
colondeath <- filter(colon, etype==2)
head(colondeath)

```

```{r}

library(tidyr)

# Replace node
colondeath <- colondeath %>%
  mutate(nodes = ifelse(nodes == 0, 1, nodes))

# Remove missing
colondeath <- colondeath %>%
  drop_na(age, sex, nodes, differ, extent, surg, node4, obstruct, perfor, adhere, rx)

head(colondeath)

```


```{r}

library(dplyr)

summary_table <- colondeath %>%
  group_by(rx) %>%
  summarize(
    Male_n = sum(sex == 1, na.rm = TRUE),
    Male_Percent = round(mean(sex == 1, na.rm = TRUE) * 100, 2),
    Female_n = sum(sex == 0, na.rm = TRUE),
    Female_Percent = round(mean(sex == 0, na.rm = TRUE) * 100, 2),
    Age_Mean_SD = paste0(round(mean(age, na.rm = TRUE), 2), " (", round(sd(age, na.rm = TRUE), 2), ")"),
    Obstruct_Yes_Percent = round(mean(obstruct == 1, na.rm = TRUE) * 100, 2),
    Perfor_Yes_Percent = round(mean(perfor == 1, na.rm = TRUE) * 100, 2),
    Adhere_Yes_Percent = round(mean(adhere == 1, na.rm = TRUE) * 100, 2),
    Nodes_Mean_SD = paste0(round(mean(nodes, na.rm = TRUE), 2), " (", round(sd(nodes, na.rm = TRUE), 2), ")"),
    Time_Mean_SD = paste0(round(mean(time, na.rm = TRUE), 2), " (", round(sd(time, na.rm = TRUE), 2), ")"),
    Differ_Well_Percent = round(mean(differ == 1, na.rm = TRUE) * 100, 2),
    Differ_Moderate_Percent = round(mean(differ == 2, na.rm = TRUE) * 100, 2),
    Differ_Poor_Percent = round(mean(differ == 3, na.rm = TRUE) * 100, 2),
    Extent_Submucosa_Percent = round(mean(extent == 1, na.rm = TRUE) * 100, 2),
    Extent_Muscle_Percent = round(mean(extent == 2, na.rm = TRUE) * 100, 2),
    Extent_Serosa_Percent = round(mean(extent == 3, na.rm = TRUE) * 100, 2),
    Extent_Contiguous_Percent = round(mean(extent == 4, na.rm = TRUE) * 100, 2),
    Surg_Short_Percent = round(mean(surg == 0, na.rm = TRUE) * 100, 2),
    Surg_Long_Percent = round(mean(surg == 1, na.rm = TRUE) * 100, 2),
    Node4_Yes_Percent = round(mean(node4 == 1, na.rm = TRUE) * 100, 2)
  ) %>%
  ungroup()

# Display the summary table
print(summary_table)

```

```{r}

library(dplyr)
library(tidyr)

# Convert all variables 
summary_table_fixed <- summary_table %>%
  mutate(across(-rx, as.character))

final_table <- summary_table_fixed %>%
  pivot_longer(cols = -rx, names_to = "Variable", values_to = "Value") %>%
  pivot_wider(names_from = rx, values_from = Value)

print(final_table)


```


Kaplan-Meier Survival Analysis

```{r}

library(survival)
library(survminer)
library(ggplot2)

surv_obj <- Surv(time = colondeath$time, event = colondeath$status)

# KM plot for rx
fit_rx <- survfit(surv_obj ~ rx, data = colondeath)
km_rx <- ggsurvplot(fit_rx,
                    data = colondeath,
                    pval = TRUE,
                    risk.table = TRUE,
                    risk.table.height = 0.25,
                    title = "Kaplan-Meier Curves by Treatment",
                    xlab = "Time (days)",
                    ylab = "Survival probability",
                    conf.int = TRUE,
                    legend.title = "Treatment",
                    ggtheme = theme_bw())

# KM plot for sex
fit_sex <- survfit(surv_obj ~ sex, data = colondeath)
km_sex <- ggsurvplot(fit_sex,
                     data = colondeath,
                     pval = TRUE,
                     risk.table = TRUE,
                     risk.table.height = 0.25,
                     title = "Kaplan-Meier Curves by Sex",
                     xlab = "Time (days)",
                     ylab = "Survival probability",
                     conf.int = TRUE,
                     legend.title = "Sex",
                     ggtheme = theme_bw())

# KM plot for diff
fit_differ <- survfit(surv_obj ~ differ, data = colondeath)
km_differ <- ggsurvplot(fit_differ,
                        data = colondeath,
                        pval = TRUE,
                        risk.table = TRUE,
                        risk.table.height = 0.25,
                        title = "Kaplan-Meier Curves by Tumor Differentiation",
                        xlab = "Time (days)",
                        ylab = "Survival probability",
                        conf.int = TRUE,
                        legend.title = "Differentiation",
                        ggtheme = theme_bw())

# KM plot for extent
fit_extent <- survfit(surv_obj ~ extent, data = colondeath)
km_extent <- ggsurvplot(fit_extent,
                        data = colondeath,
                        pval = TRUE,
                        risk.table = TRUE,
                        risk.table.height = 0.25,
                        title = "Kaplan-Meier Curves by Extent",
                        xlab = "Time (days)",
                        ylab = "Survival probability",
                        conf.int = TRUE,
                        legend.title = "Extent",
                        ggtheme = theme_bw())

# KM plot for node4
fit_node4 <- survfit(surv_obj ~ node4, data = colondeath)
km_node4 <- ggsurvplot(fit_node4,
                       data = colondeath,
                       pval = TRUE,
                       risk.table = TRUE,
                       risk.table.height = 0.25,
                       title = "Kaplan-Meier Curves by Lymph Node Status",
                       xlab = "Time (days)",
                       ylab = "Survival probability",
                       conf.int = TRUE,
                       legend.title = "Node4",
                       ggtheme = theme_bw())

print(km_rx)
print(km_sex)
print(km_differ)
print(km_extent)
print(km_node4)

```

Log-rank Tests

```{r}

library(survival)
library(survminer)
library(ggplot2)

surv_obj <- Surv(time = colondeath$time, event = colondeath$status)

# Log-rank tests rx
lr_rx <- survdiff(surv_obj ~ rx, data = colondeath)
print("Log-rank test results for Treatment groups:")
print(lr_rx)
cat("\nP-value:", 1 - pchisq(lr_rx$chisq, df = length(lr_rx$n) - 1), "\n\n")

# Sex
lr_sex <- survdiff(surv_obj ~ sex, data = colondeath)
print("Log-rank test results for Sex:")
print(lr_sex)
cat("\nP-value:", 1 - pchisq(lr_sex$chisq, df = length(lr_sex$n) - 1), "\n\n")

# Differentiation
lr_differ <- survdiff(surv_obj ~ differ, data = colondeath)
print("Log-rank test results for Tumor Differentiation:")
print(lr_differ)
cat("\nP-value:", 1 - pchisq(lr_differ$chisq, df = length(lr_differ$n) - 1), "\n\n")

# Extent 
lr_extent <- survdiff(surv_obj ~ extent, data = colondeath)
print("Log-rank test results for Extent of Spread:")
print(lr_extent)
cat("\nP-value:", 1 - pchisq(lr_extent$chisq, df = length(lr_extent$n) - 1), "\n\n")

# node4
lr_node4 <- survdiff(surv_obj ~ node4, data = colondeath)
print("Log-rank test results for Lymph Node Status:")
print(lr_node4)
cat("\nP-value:", 1 - pchisq(lr_node4$chisq, df = length(lr_node4$n) - 1), "\n\n")

log_rank_summary <- data.frame(
  Variable = c("Treatment (rx)", "Sex", "Differentiation", "Extent", "Node4"),
  Chi_Square = c(lr_rx$chisq, lr_sex$chisq, lr_differ$chisq, lr_extent$chisq, lr_node4$chisq),
  Degrees_of_Freedom = c(length(lr_rx$n) - 1, 
                        length(lr_sex$n) - 1,
                        length(lr_differ$n) - 1,
                        length(lr_extent$n) - 1,
                        length(lr_node4$n) - 1),
  P_Value = c(1 - pchisq(lr_rx$chisq, df = length(lr_rx$n) - 1),
              1 - pchisq(lr_sex$chisq, df = length(lr_sex$n) - 1),
              1 - pchisq(lr_differ$chisq, df = length(lr_differ$n) - 1),
              1 - pchisq(lr_extent$chisq, df = length(lr_extent$n) - 1),
              1 - pchisq(lr_node4$chisq, df = length(lr_node4$n) - 1))
)

# P-values 
log_rank_summary$P_Value <- format.pval(log_rank_summary$P_Value, digits = 4)

print(log_rank_summary)
```


Unadjusted Cox Proportional Hazards Model

```{r}

library(survival)

# unadjusted Cox model for treatment
cox_treatment <- coxph(Surv(time, status) ~ rx, data = colondeath)
summary(cox_treatment)

# hazard ratios and confidence intervals
cox_hr <- exp(cbind(coef(cox_treatment), confint(cox_treatment)))
colnames(cox_hr) <- c("Hazard Ratio", "Lower 95% CI", "Upper 95% CI")

print("Hazard Ratios with 95% Confidence Intervals:")
print(cox_hr)

# proportional hazards assumption
ph_test <- cox.zph(cox_treatment)
print("\nProportional Hazards Assumption Test:")
print(ph_test)

plot(ph_test)
```



Adjusted Cox Proportional Hazards Model

```{r}

library(survival)

# adjusted Cox model with all confounders
cox_adjusted <- coxph(Surv(time, status) ~ rx + sex + age + obstruct + perfor + 
                     adhere + nodes + differ + extent + surg + node4,
                     data = colondeath)

summary(cox_adjusted)

# hazard ratios and confidence intervals
cox_hr <- exp(cbind(coef(cox_adjusted), confint(cox_adjusted)))
colnames(cox_hr) <- c("Hazard Ratio", "Lower 95% CI", "Upper 95% CI")

print("Hazard Ratios with 95% Confidence Intervals:")
print(cox_hr)

# proportional hazards assumption
ph_test <- cox.zph(cox_adjusted)
print("\nProportional Hazards Assumption Test:")
print(ph_test)

library(forestplot)

coef_names <- names(coef(cox_adjusted))

hr_data <- data.frame(
  Variable = coef_names,
  HR = exp(coef(cox_adjusted)),
  Lower = exp(confint(cox_adjusted))[,1],
  Upper = exp(confint(cox_adjusted))[,2],
  P = summary(cox_adjusted)$coefficients[,5]
)

# forest plot
forestplot(
  labeltext = hr_data$Variable,
  mean = hr_data$HR,
  lower = hr_data$Lower,
  upper = hr_data$Upper,
  title = "Adjusted Hazard Ratios with 95% CI",
  xlab = "Hazard Ratio",
  zero = 1,
  boxsize = 0.2,
  col = fpColors(box = "royalblue",
                line = "darkblue",
                summary = "royalblue")
)

# likelihood ratio test
null_model <- coxph(Surv(time, status) ~ 1, data = colondeath)
lr_test <- anova(null_model, cox_adjusted)
print("\nLikelihood Ratio Test:")
print(lr_test)

conc <- concordance(cox_adjusted)
print("\nConcordance Index:")
print(conc)
```

Test proportional hazard assumptions

```{r}

library(survival)

cox_adjusted <- coxph(Surv(time, status) ~ rx + sex + age + obstruct + perfor + 
                     adhere + nodes + differ + extent + surg + node4,
                     data = colondeath)

# proportional hazards assumption
ph_test <- cox.zph(cox_adjusted)
print("Proportional Hazards Test Results:")
print(ph_test)

model_summary <- summary(cox_adjusted)

results_table <- data.frame(
  Variable = names(coef(cox_adjusted)),
  Hazard_Ratio = round(exp(coef(cox_adjusted)), 3),
  Lower_CI = round(exp(confint(cox_adjusted))[,1], 3),
  Upper_CI = round(exp(confint(cox_adjusted))[,2], 3),
  P_value = round(model_summary$coefficients[,5], 4)
)

results_table$CI_95 <- paste0("(", results_table$Lower_CI, " - ", results_table$Upper_CI, ")")

# table
final_table <- data.frame(
  Variable = results_table$Variable,
  HR = results_table$Hazard_Ratio,
  CI_95 = results_table$CI_95,
  P_value = results_table$P_value
)

print("\nCox Model Results:")
print(final_table, row.names = FALSE)

# Plot
print("\nPlotting Schoenfeld residuals...")
plot(ph_test)
print("\nGlobal Test of Proportional Hazards Assumption:")
print(ph_test$table["GLOBAL",])

# residuals
mart_resid <- residuals(cox_adjusted, type="martingale")
dev_resid <- residuals(cox_adjusted, type="deviance")

par(mfrow=c(2,2))

# Plots
plot(predict(cox_adjusted), mart_resid,
     xlab="Linear Predictor",
     ylab="Martingale Residuals",
     main="Martingale Residuals vs. Linear Predictor")
abline(h=0, col="red", lty=2)

plot(predict(cox_adjusted), dev_resid,
     xlab="Linear Predictor",
     ylab="Deviance Residuals",
     main="Deviance Residuals vs. Linear Predictor")
abline(h=0, col="red", lty=2)

par(mfrow=c(1,1))

```

Due to assumption violation, stratification by differ

```{r}

library(survival)
library(survminer)

# stratified Cox model with 'differ' 
cox_stratified <- coxph(Surv(time, status) ~ rx + sex + age + obstruct + perfor + 
                       adhere + nodes + extent + surg + node4 + strata(differ),
                       data = colondeath)

# Test PH assumption 
ph_test_strat <- cox.zph(cox_stratified)
print("Proportional Hazards Test Results (Stratified Model):")
print(ph_test_strat)

model_summary_strat <- summary(cox_stratified)
results_table_strat <- data.frame(
    Variable = names(coef(cox_stratified)),
    Hazard_Ratio = round(exp(coef(cox_stratified)), 3),
    Lower_CI = round(exp(confint(cox_stratified))[,1], 3),
    Upper_CI = round(exp(confint(cox_stratified))[,2], 3),
    P_value = round(model_summary_strat$coefficients[,5], 4)
)

# confidence intervals
results_table_strat$CI_95 <- paste0("(", results_table_strat$Lower_CI, 
                                   " - ", results_table_strat$Upper_CI, ")")

final_table_strat <- data.frame(
    Variable = results_table_strat$Variable,
    HR = results_table_strat$Hazard_Ratio,
    CI_95 = results_table_strat$CI_95,
    P_value = results_table_strat$P_value
)

print("\nStratified Cox Model Results:")
print(final_table_strat, row.names = FALSE)

# plots
par(mfrow=c(2,2))
plot(ph_test_strat)
mart_resid_strat <- residuals(cox_stratified, type="martingale")
dev_resid_strat <- residuals(cox_stratified, type="deviance")

plot(predict(cox_stratified), mart_resid_strat,
     xlab="Linear Predictor",
     ylab="Martingale Residuals",
     main="Martingale Residuals (Stratified Model)")
lines(lowess(predict(cox_stratified), mart_resid_strat), col="red")
abline(h=0, lty=2)

plot(predict(cox_stratified), dev_resid_strat,
     xlab="Linear Predictor",
     ylab="Deviance Residuals",
     main="Deviance Residuals (Stratified Model)")
lines(lowess(predict(cox_stratified), dev_resid_strat), col="red")
abline(h=0, lty=2)

par(mfrow=c(1,1))

# survival curves
ggsurvplot(
    survfit(cox_stratified),
    data = colondeath,
    risk.table = TRUE,
    ggtheme = theme_minimal(),
    title = "Survival Curves by Differentiation Strata"
)

# AIC
AIC_original <- AIC(cox_adjusted)
AIC_stratified <- AIC(cox_stratified)

print("\nModel Comparison (AIC):")
print(data.frame(
    Model = c("Original", "Stratified"),
    AIC = c(AIC_original, AIC_stratified)
))

# likelihood ratio test 
lrt <- anova(cox_adjusted, cox_stratified)
print("\nLikelihood Ratio Test Results:")
print(lrt)
```

Significantly improved but still has violation, can refine with obstruct.
So, adjusted cox with obstruct and differ.

```{r}

library(survival)
library(survminer)
library(forestplot)

# Fit model with 2 stratification
cox_refined <- coxph(Surv(time, status) ~ rx + sex + age + perfor + 
                    adhere + nodes + extent + surg + node4 +
                    strata(differ, obstruct),
                    data = colondeath)

# PH assumption 
ph_test_refined <- cox.zph(cox_refined)
print("Proportional Hazards Test Results (Refined Model):")
print(ph_test_refined)

model_summary_refined <- summary(cox_refined)

results_table_refined <- data.frame(
    Variable = names(coef(cox_refined)),
    Hazard_Ratio = round(exp(coef(cox_refined)), 3),
    Lower_CI = round(exp(confint(cox_refined))[,1], 3),
    Upper_CI = round(exp(confint(cox_refined))[,2], 3),
    P_value = round(model_summary_refined$coefficients[,5], 4)
)

# confidence intervals
results_table_refined$CI_95 <- paste0("(", results_table_refined$Lower_CI, 
                                   " - ", results_table_refined$Upper_CI, ")")

final_table_refined <- data.frame(
    Variable = results_table_refined$Variable,
    HR = results_table_refined$Hazard_Ratio,
    CI_95 = results_table_refined$CI_95,
    P_value = results_table_refined$P_value
)

print("\nRefined Cox Model Results:")
print(final_table_refined, row.names = FALSE)

par(mfrow=c(2,2))

# Plots
plot(ph_test_refined)

mart_resid_refined <- residuals(cox_refined, type="martingale")
dev_resid_refined <- residuals(cox_refined, type="deviance")

plot(predict(cox_refined), mart_resid_refined,
     xlab="Linear Predictor",
     ylab="Martingale Residuals",
     main="Martingale Residuals (Refined Model)")
lines(lowess(predict(cox_refined), mart_resid_refined), col="red")
abline(h=0, lty=2)

plot(predict(cox_refined), dev_resid_refined,
     xlab="Linear Predictor",
     ylab="Deviance Residuals",
     main="Deviance Residuals (Refined Model)")
lines(lowess(predict(cox_refined), dev_resid_refined), col="red")
abline(h=0, lty=2)

par(mfrow=c(1,1))

# AIC
AIC_original <- AIC(cox_adjusted)
AIC_stratified <- AIC(cox_stratified)
AIC_refined <- AIC(cox_refined)

print("\nModel Comparison (AIC):")
print(data.frame(
    Model = c("Original", "Stratified", "Refined"),
    AIC = c(AIC_original, AIC_stratified, AIC_refined)
))

# likelihood ratio test 
lrt <- anova(cox_adjusted, cox_stratified, cox_refined)
print("\nLikelihood Ratio Test Results:")
print(lrt)

ggsurvplot(
  survfit(cox_refined),
  data = colondeath,
  risk.table = TRUE,
  conf.int = FALSE,
  ggtheme = theme_minimal(),
  title = "Survival Curves by Strata",
  font.main = c(14, "bold"),
  font.x = 12,
  font.y = 12,
  font.tickslab = 10,
  risk.table.height = 0.25,
  risk.table.fontsize = 2,    
  risk.table.title = "Number at Risk",
  risk.table.y.text = FALSE,
  tables.theme = theme_classic()
)

```



LASSO Penalized Regression


```{r}
library(glmnet)
library(dplyr)
library(survival)

surv_obj_strata <- with(colondeath, 
                        glmnet::stratifySurv(Surv(time, status),
                                           strata = interaction(obstruct, differ)))

# Scale continuous variables
covs <- model.matrix(~ rx + age + sex + perfor + adhere + 
                    nodes + extent + surg,
                    data = colondeath)[,-1] %>%
        as.data.frame() %>%
        mutate_at(vars(age, nodes), scale) %>%
        as.matrix()

set.seed(123)

# Fit LASSO model
lasso_fit <- suppressWarnings(
    cv.glmnet(x = covs,
              y = surv_obj_strata,
              family = "cox",
              alpha = 1)
)

# Plot 
plot(lasso_fit)
title("Cross-validation Plot for LASSO")

# minimum lambda and 1SE lambda
coef_min <- coef(lasso_fit, s = "lambda.min")
coef_1se <- coef(lasso_fit, s = "lambda.1se")

format_coef <- function(coef_matrix) {
    coef_df <- data.frame(
        Variable = rownames(coef_matrix),
        Coefficient = as.vector(coef_matrix)
    ) %>%
    filter(Coefficient != 0) %>%
    arrange(desc(abs(Coefficient)))
    
    return(coef_df)
}

# Outputs
print("Variables selected using minimum lambda:")
print(format_coef(coef_min))
print("\nVariables selected using lambda.1se (more conservative):")
print(format_coef(coef_1se))
print("\nOptimal lambda values:")
print(paste("Minimum lambda:", round(lasso_fit$lambda.min, 4)))
print(paste("Lambda 1SE:", round(lasso_fit$lambda.1se, 4)))

# final Cox model with using lambda.1se
selected_vars <- rownames(coef_1se)[which(coef_1se != 0)]
selected_vars <- selected_vars[selected_vars != "(Intercept)"]
formula_str <- paste("Surv(time, status) ~", 
                    paste(selected_vars, collapse = " + "))
final_cox <- coxph(as.formula(formula_str), data = colondeath)
print("\nFinal Cox Model with LASSO-selected variables:")
print(summary(final_cox))

```


Model Comparison and Final Model Selection

```{r}

library(survival)
library(glmnet)
library(dplyr)

# all three models
cox_unadj <- cox_treatment
cox_full <- cox_refined
cox_lasso <- final_cox  

# comparison table
model_comparison <- data.frame(
    Model = c("Unadjusted", "Full", "LASSO"),
    AIC = c(AIC(cox_unadj), AIC(cox_full), AIC(cox_lasso)),
    BIC = c(BIC(cox_unadj), BIC(cox_full), BIC(cox_lasso)),
    C_index = c(summary(cox_unadj)$concordance[1],
                summary(cox_full)$concordance[1],
                summary(cox_lasso)$concordance[1]),
    Parameters = c(length(coef(cox_unadj)),
                  length(coef(cox_full)),
                  length(coef(cox_lasso))),
    Events = c(cox_unadj$nevent, cox_full$nevent, cox_lasso$nevent),
    Sample_size = c(cox_unadj$n, cox_full$n, cox_lasso$n)
) %>%
mutate(across(where(is.numeric), round, 3))

# Likelihood ratio tests
lr_tests <- list(
    full_vs_unadj = anova(cox_unadj, cox_full),
    lasso_vs_unadj = anova(cox_unadj, cox_lasso),
    full_vs_lasso = anova(cox_lasso, cox_full)
)

# discrimination
c_improvement <- data.frame(
    Model = c("Full", "LASSO"),
    Improvement = c(
        (model_comparison$C_index[2] - model_comparison$C_index[1]) / 
            (1 - model_comparison$C_index[1]) * 100,
        (model_comparison$C_index[3] - model_comparison$C_index[1]) / 
            (1 - model_comparison$C_index[1]) * 100
    )
) %>%
mutate(Improvement = round(Improvement, 1))

# results
cat("\nModel Comparison Summary:\n")
print(model_comparison)

cat("\nRelative Improvement in Discrimination (% vs Unadjusted):\n")
print(c_improvement)

cat("\nLikelihood Ratio Tests:\n")
cat("\nFull vs Unadjusted:\n")
print(lr_tests$full_vs_unadj)
cat("\nLASSO vs Unadjusted:\n")
print(lr_tests$lasso_vs_unadj)
cat("\nFull vs LASSO:\n")
print(lr_tests$full_vs_lasso)
cat("\nModel Selection Summary:\n")
best_model <- data.frame(
    Criterion = c("AIC", "BIC", "C-index"),
    Best_Model = c(
        model_comparison$Model[which.min(model_comparison$AIC)],
        model_comparison$Model[which.min(model_comparison$BIC)],
        model_comparison$Model[which.max(model_comparison$C_index)]
    )
)
print(best_model)
```



